---
title: "Mangrove Exploration -- Misc. Exploration"
author: Nicholas Ko
output: 
 html_document:
    toc : TRUE
    toc_float: TRUE
    number_sections: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Background

This Rmd file contains miscellaneous data exploration and plots that I did in my spare time. 

**load libraries**
```{r}
library(tidyverse)
library(ggplot2)
library(heatmaply)
library(wesanderson)
library(ggpubr)
```


**load data**
```{r}
species <- read_csv("processed_data/species.csv")
tree_probs <- read_csv("processed_data/tree_probs_new.csv")
```


# Misc. Exploration: 

**Load background objects**

The code for these objects is taken from the file `mangrove_exploration.Rmd`

```{r}
sp_rank <- species %>%
  rename(species = spp) %>%
  select(species, est_median) %>%
  filter(!species %in% c("LURA", "SCHY")) %>%
  #create column with ranked est_median
  mutate(rank = rank(est_median)) %>%
  #arrange in ascending order
  arrange(rank)


sp_abundance_2 <- tree_probs %>% 
  group_by(treeid, plot, spp_orig) %>%
  summarise(n_trees = n()) %>%
  select(-n_trees) %>%
  #remove unwanted species, low count and random species
  filter(!spp_orig %in% c("LURA","SCHY", "random")) %>%
  group_by(plot, spp_orig) %>% 
  summarise(n_tree = n()) %>%
  #pivot
  pivot_wider(
  names_from = spp_orig, 
  values_from = n_tree,
  values_fill = 0
) %>%
  #calculate sum for each row
  rowwise() %>%
  #sum total number of trees per plot
  mutate(total_trees = sum(c_across(AVOF:EXAG))) %>%
  #repivot, now with 0 added to values fill
  pivot_longer(
  cols = !c(plot,total_trees),
  names_to = "species",
  values_to = "n_trees") %>%
  #add elevation ranking
  left_join(sp_rank) %>%
  ungroup()
```

## Species Heat map

Rank species according to elevation

```{r}
elevation_rank <- species %>%
 # select(species, nr_trees) %>%
  left_join(sp_abundance_2) %>%
  filter(!species %in% c("LURA", "SCHY")) %>%
  rename(trees_per_species = nr_trees) %>%
  relocate(plot, .before = species) %>%
  arrange(plot) %>%
  mutate(species_contribution = 100*(n_trees/trees_per_species)) %>%
  mutate(species_contribution = round(species_contribution, digits = 1)) %>%
  relocate(plot, .before = nr_plots) %>%
  relocate(n_trees, .after = nr_plots) %>%
  relocate(species_contribution, .after = trees_per_species) %>%
  relocate(rank, .after = spp) %>%
  arrange(spp) %>%
  arrange(rank) %>%
  select(-species) %>%
  ungroup()

#species_contribution col: proportion of trees represented by a species at plot level
```

First 20 rows of `elevation_rank` data frame

```{r fig.height = 7, fig.width = 15, include = FALSE}
elev_rank_ggtt <- elevation_rank
  
elev_rank_ggtt[,-1] <- round(elev_rank_ggtt[,-1],1)

ggtexttable(elev_rank_ggtt[1:20,], 
            theme = ttheme("light",
                           base_size = 10.5, 
                           padding = unit(c(5, 4),
                                          "mm"))) %>%
  tab_add_title(text = "elevation_rank data frame \n (first 20 rows)")

rm(elev_rank_ggtt)
```


### data frame

```{r}
#heatmap data frame
heatmap_df <- tree_probs %>%
  filter(!spp_orig %in% c("LURA", "SCHY", "random")) %>%
  group_by(treeid,
           spp_orig,
           elev_bin) %>%
  #x = number of treeid replicates, will remove later
  summarise(x = n()) %>%
  arrange(treeid) %>%
  select(treeid, spp_orig, elev_bin) %>%
  group_by(spp_orig, elev_bin) %>%
  summarise(tree_per_bin = n()) %>%
  rename(species = spp_orig) %>%
  ungroup()

```

### ggplot2

```{r}
#species heatmap with ggplot2
ggplot(heatmap_df, aes(x = species, 
                                 y = factor(elev_bin), 
                                 fill = tree_per_bin)) +
  #use geom_tile for heatmap, look into heatmap.2() function
  geom_tile() +
  coord_flip() + 
  labs(x = "Species",
       #issue: too many elevation bins, 431 of them
       y = "Elevation Bin",
       fill = "Species Abundance") + 
  scale_fill_gradient(low="white", high="red") + 
  theme_classic()

#ggplot2 heat map ver 2.0
pal <- wes_palette("Zissou1", 100, type = "continuous")
heatmap_v2 <- ggplot(heatmap_df, aes(x = elev_bin, y = species, fill = tree_per_bin)) +
  geom_tile() + 
  scale_fill_gradientn(colours = pal) + 
  scale_x_discrete(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0)) + 
  coord_equal() +
  theme_classic()
```

```{r export ggplot2 heatmap, eval=FALSE, include=FALSE}
#export v2.0 to pdf
ggexport(heatmap_v2, filename="heatmap_v2.pdf", width = 250, height = 10)
```


Using heatmaply
```{r eval=FALSE, include=FALSE}
heatmap_df2 <- heatmap_df %>% 
  pivot_wider(names_from = elev_bin, 
              values_from = tree_per_bin,
              values_fill = 0)

#species heat map with heatmaply
heatmaply(heatmap_df2, 
        dendrogram = "none",
        col = heat,
        xlab = "Elevation bin", 
        ylab = "Species")
```




```{r junk code, eval=FALSE, include=FALSE}
+
  labs(x = "Plot", 
       y = "Number of trees",
       fill = "Species") + 
  coord_flip() +
  theme(legend.position = "bottom",
        legend.box = "horizontal") +
  guides(fill=guide_legend(ncol = 16,byrow=TRUE)) 
```


## Plot vs elevation rank (w/ abundance)

```{r}
#exploratory bar plot of distribution of species per plot
sp_plot <- ggplot(sp_abundance_2, aes(x = factor(plot), 
                                      y = n_trees, 
                                      #code below orders the plant species based on the elevation rank
                                      fill = fct_reorder(species, rank, .desc = FALSE))) +
  geom_col() +
  labs(x = "Plot", 
       y = "Number of trees",
       fill = "Species") + 
  coord_flip() +
  theme(legend.position = "bottom",
        legend.box = "horizontal") +
  guides(fill=guide_legend(ncol = 16,byrow=TRUE)) 

sp_plot

```



```{r fig.height = 20, fig.width = 20}
#use species data frame which includes median elevation component

order_plot <- ggplot(species[-c(1,2),], aes(x = est_median, y = reorder(spp, est_median))) +
  geom_point(size = 3, col = "blue") +  # Use a larger dot
  theme_bw() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.major.y = element_line(colour = "grey60", linetype = "dashed"),
    axis.text.x = element_text(size = 20),
    axis.text.y = element_text(size = 20),
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(size = 20),
    plot.title = element_text(size = 20),
  ) +
  labs(x = "Estimated Median Elevation",
       y = "Species of Tree") +
  ggtitle("Order of trees along elevation gradient") +
  scale_x_continuous(breaks=seq(0,400,10))

order_plot

```

### Export plot

```{r eval = FALSE}
multi.page <- ggarrange(order_plot,
                        sp_plot,
                        #specify number of rows and columns
                        nrow=2, ncol=1) 
multi.page[[1]] 

ggexport(multi.page, filename="elevation_plot.pdf", width = 16, height = 20)

ggexport(order_plot, filename="elev_by_sp.pdf", width = 16, height = 12)
```

